Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-01-30 PAPER.tex
edited/cleaned up Danny's revisions in 3.1.1

In Section \ref{sec:SiglossOverview}, we showed how signal loss arises when weighting data using information from the data

itself. Here we describe a methodology for estimating the amount of signal loss caused by a particular power spectrum estimator when applied to a particular data set. The exact signal loss ``correction factor" for a particular data set will depend on the specific realizations of the signals present in the data and is not something we can directly compute. 
In this work, we inject simulated cosmological signals into our data and test the recovery of those signals in order to statistically quantify 
the amount of signal loss accompanying a weighting scheme used in OQE power spectrum estimation (an approach also taken by \citet{masui_et_al2013}).  
The results of such an injection scheme can be interpreted in multiple ways.  In this paper we describe two. This section describes one and in Appendix \ref{sec:Pr_appendix} we present an alternate interpretation.
These two methods yield very similar results, but fundamentally are answering subtly distinct questions.  
We also highlight major differences from the signal loss 
computation used in A15, which previously underestimated losses.

\subsubsection{Signal Loss Methodology} 
\label{sec:siglossmethod}
The gist of our method is to inject a eor-like signal into the data and then measure the amount recovered, with the difference being accounted as signal loss.  To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ 
cm signal visibilities. Here we use the same method as in A15, where mock Gaussian noise visibilities (mock EoR signals) 
are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.\footnote{One 
specific change from A15 is that we add this simulated signal into the analysis pipeline before the final fringe-rate filter is 
applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased 
estimate of signal loss, %(by a factor of $\sim$$10$), 
likely due to the use of the fringe-rate filter as a simulator. However, this pipeline difference, while significant, is not the dominant reason why signal loss was under-estimated in A15 (the dominant reason is explained in the main text in Section \ref{sec:Illustration}).}

@@

\begin{eqnarray}
\label{eq:sigloss}
P_{\rm out,\alpha} &\equiv& \widehat{\textbf{p}}_{r,\alpha}-\widehat{\textbf{p}}_{x,\alpha} \nonumber \\
&=& M^{\alpha}_{r}\textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r} - M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x},
\end{eqnarray}
where, for illustrative purposes and notational simplicity, we have written these equations with diagonal normalization matrices $M$, even though for our numerical results we choose a non-diagonal matrix $\mathbf{M}$ as in Equation \eqref{eq:phat}. $P_{\rm in}$ represents the unweighted power spectrum of $\textbf{e}$, our simulated EoR signal, 
$\widehat{\textbf{p}}_{x}$ is the power spectrum of the data alone, $\widehat{\textbf{p}}_{r}$ is the power spectrum of the
data plus injection, and $P_{\rm out}$ is the difference between $\widehat{\textbf{p}}_{r}$ and $\widehat{\textbf{p}}_{x}$.  In the limit where no significant cross correlations between the data and the injected signals are found, this will recover the amount of injected power spectrum remaining after the covariance weighting operation, ie the amount of background EoR power $P_{\rm eor}$ not removed by the foreground downweighting procedure. In other words, we make the anzatz that

\begin{equation}
P_{\rm out} = P_{\rm eor} \label{eqn:anzatz},
\end{equation}
folding in our definition from Equation \ref{eq:sigloss}, this definition implies that we are seeking the injected 
signal amplitude which causes a power doubling ($\phat_r = 2 \phat_x$).

@@

\begin{eqnarray}
\label{eq:crossterm}
P_{\rm out,\alpha} &=& M^{\alpha}_{r}(\textbf{x}+\textbf{e})^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}(\textbf{x}+\textbf{e}) - 
M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x} \nonumber \\
&=& M^{\alpha}_{a}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} + M^{\alpha}_{b}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}
^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&+& M^{\alpha}_{c}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} + M^{\alpha}_{d}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} \nonumber \\
&-& M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}

Assuming R$_{r}$ is symmetric the two cross terms can be summed together as

\begin{eqnarray}
\label{eq:crossterm}
P_{\rm out,\alpha} &= &  M^{\alpha}_{a}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} \\
&+& M^{\alpha}_{b}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}
^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&+& 2 M^{\alpha}_{c}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} \\
&-& M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}.

Each of these terms is plotted in Figure \ref{fig:sigloss_terms} for two cases: inverse covariance weighting ($\textbf{R}_{r} = \widehat{\textbf{C}}_{r}^{-1}$) and unweighted ($\textbf{R}_{r} = \textbf{I}$).   Let us examine the behavior of this equation in the regimes of very weak and very strong injected signal.

{\bf Small injection}
The first and the fourth term are very similar, differing only in their covariance matrices. When the injected signal 
is very small compared to the data, their difference can be neglected.  In this limit we might imagine that the 
cross correlation should also be small enabling us to recover just the EoR term, however examining the cross-
term plotted in Figure  \ref{fig:sigloss_terms} we see that cross terms are still large enough to have a significant 
impact. This is trivially demonstrated in the identity case where we even at small injects, large cross terms are 
evident.

{\bf Large injection}
When the injected signal is much larger than the data-only power spectrum, the data-only components can 
be neglected as many orders of magnitude smaller. As we can see from Figure  \ref{fig:sigloss_terms}, the cross terms are also small in comparison.  Here only do we recover the ``loss'' injected simulated eor 
signal as planned.

{\bf In between}
When the injected signal is of a similar amplitude to the data by itself, the picture becomes murky.  We see that 
the weighted injected power spectrum component mirrors the input power indicating little loss, eventually 
departing from unity when the injected amplitude is well above the level of the data power spectrum. However 
the cross terms have nearly the same amplitude, but a negative sign. The sum of the two is substantially lower and substantially drives down the $P_{\rm out}$ estimator.

@@

The source of the strong negative cross term is not immediately obvious, however it is an explainable effect. 
When $\textbf{R}_{r}$
is taken to be $\widehat{\textbf{C}}_{r}^{-1}$, the second term of Equation \eqref{eq:pout_expand} is a cross-correlation between $\widehat{\textbf{C}}_{r}^{-1}\textbf{x}$ and
$\widehat{\textbf{C}}_{r}^{-1}\textbf{e}$. As shown in \citet{switzer_et_al2015}, this cross-correlation term is non-zero, and in fact negative in expectation. 
This negative cross-term power arises from a coupling between the inverse of 
$\widehat{\textbf{C}}_{r}$ and $\mathbf{x}$. 
Intuitively, we can see this by expanding the empirical covariance of $\textbf{r}=\textbf{x}+\textbf{e}$ in the
$|\textbf{x}|\ll|\textbf{e}|$ regime:

@@

\noindent where we can neglect the first term because $\textbf{x}$ is small.  Without loss of generality, we will assume
an eigenbasis of $\textbf{e}$, so that $\langle 
\textbf{ee}^{\dagger} \rangle_{t}$ is diagonal. 
The middle 
two terms, however, can have power in their off-diagonal terms due to the fact that, when averaging over a finite
ensemble, $\langle\textbf{xe}^\dagger\rangle_t$ is not zero.  As shown in Appendix C of \citet{parsons_et_al2014}, to leading order the
inversion of a matrix like $\widehat{\textbf{C}}_{r}$ that is diagonal-dominant (from $\langle 
\textbf{ee}^{\dagger} \rangle_{t}$) with smaller
off-diagonal terms results in a new diagonal-dominant matrix with negative off-diagonal terms. These off-diagonal
terms depend on both $\textbf{x}$ and $\textbf{e}$. When $\widehat{\textbf{C}}^{-1}_{r}$ is multiplied into $\textbf{x}$,
the result is a vector that is similar to $\textbf{x}$ but
contains a residual correlation to $\textbf{e}$ from the off-diagonal components of $\widehat{\textbf{C}}^{-1}_{r}$. The
correlation is negative because the product $\widehat{\textbf{C}}_r^{-1}\textbf{x}$ effectively squares the $\textbf{x}$-dependence
of the off-diagonal terms in $\widehat{\textbf{C}}^{-1}_{r}$ while retaining the negative sign that arose from the inversion
of a diagonal-dominant matrix.
% XXX [ARP: we may want to write this all out explicitly in an appendix]

The correlation between $\widehat{\textbf{C}}_r^{-1}\textbf{e}$ and $\widehat{\textbf{C}}_r^{-1}\textbf{x}$ means that we cannot compute signal loss using a signal-only 
simulation, which would yield greater values for $P_{\rm out}$ and thereby underestimate signal loss. This argument is also made in \citet{switzer_et_al2015} in regards to estimating loss associated with foreground cleaning of Great Bank Telescope (GBT) data. Therefore, in our revised 
signal loss computation we use the full quantity for $P_{\rm out}$ as defined in Equation \eqref{eq:sigloss}, which subtracts the 
weighted power spectrum of the data from the weighted power spectrum of data plus EoR.

>>>>

In Section \ref{sec:SiglossOverview}, we showed how signal loss arises when weighting data using information from the data itself. Here we describe a methodology for estimating the amount of signal loss caused by a particular power spectrum estimator when applied to a particular dataset. The exact amount of signal loss will depend on the specific realizations of the signals present in the data and is not something we can directly compute. In this work, as in A15, we inject simulated cosmological signals into our data and test the recovery of those signals (an approach also taken by \citet{masui_et_al2013}). As we will see, correlations between the injected signals and the data are significant complicating factors which were previously not taken into account.

The results of such an injection scheme can be interpreted in multiple ways. In this paper we describe two --- this section describes one and in Appendix \ref{sec:Pr_appendix} we present an alternate interpretation. These two methods yield very similar results, but fundamentally are answering subtly distinct questions. We also highlight major differences from the signal loss computation used in A15, which previously underestimated losses.

\subsubsection{Signal Loss Methodology} 
\label{sec:siglossmethod}
In short, our method consists of injecting an EoR-like signal into the data and then measuring the amount recovered, with the difference being accounted as signal loss.  To capture the full statistical likelihood of signal loss, one requires a quick way to generate many realizations of simulated $21$ 
cm signal visibilities. Here we use the same method as in A15, where mock Gaussian noise visibilities (mock EoR signals) 
are filtered in fringe-rate space to retain only ``sky-like" time-modes. This signal is then added to the data.\footnote{One 
specific change from A15 is that we add this simulated signal into the analysis pipeline before the final fringe-rate filter is 
applied to the data. Previously, the addition was done after that final fringe-rate filter step.  This change results in an increased 
estimate of signal loss, %(by a factor of $\sim$$10$), 
likely due to the use of the fringe-rate filter as a simulator. However, this pipeline difference, while significant, is not the dominant reason why signal loss was under-estimated in A15 (the dominant reason is explained in the main text in Section \ref{sec:siglossmethod}).}

@@

\begin{eqnarray}
\label{eq:sigloss}
P_{\rm out,\alpha} &\equiv& \widehat{\textbf{p}}_{r,\alpha}-\widehat{\textbf{p}}_{x,\alpha} \nonumber \\
&=& M^{\alpha}_{r}\textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r} - M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x},
\end{eqnarray}
where, for illustrative purposes and notational simplicity, we have written these equations with diagonal normalization matrices $M$, even though for our numerical results we choose a non-diagonal matrix $\mathbf{M}$ as in Equation \eqref{eq:phat}. $P_{\rm in}$ represents the unweighted power spectrum of $\textbf{e}$, our simulated EoR signal, 
$\widehat{\textbf{p}}_{x}$ is the power spectrum of the data alone, $\widehat{\textbf{p}}_{r}$ is the power spectrum of the
data plus injection, and $P_{\rm out}$ is the difference between $\widehat{\textbf{p}}_{r}$ and $\widehat{\textbf{p}}_{x}$.  In the limit where no significant cross correlations between the data and the injected signals are found, this will recover the amount of injected power spectrum remaining after the covariance weighting operation (i.e. the amount of background EoR power, $P_{\rm eor}$, not removed by the foreground down-weighting procedure). In other words, we make the anzatz that

\begin{equation}
P_{\rm out} = P_{\rm eor} \label{eqn:anzatz},
\end{equation}
and folding in our definition from Equation \eqref{eq:sigloss}, this ansatz implies that we are seeking the injected 
signal amplitude which causes a power doubling ($\phat_r = 2 \phat_x$). Expanding out Equation \eqref{eq:sigloss}, $P_{\rm out}$ becomes:

@@

\begin{eqnarray}
\label{eq:crossterm}
P_{\rm out,\alpha} &=& M^{\alpha}_{r}(\textbf{x}+\textbf{e})^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}(\textbf{x}+\textbf{e}) - 
M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x} \nonumber \\
&=& M^{\alpha}_{a}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} + M^{\alpha}_{b}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}
^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&+& M^{\alpha}_{c}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} + M^{\alpha}_{d}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} \nonumber \\
&-& M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}
Assuming \textbf{R}$_{r}$ is symmetric, the two cross-terms can be summed together as:

\begin{eqnarray}
\label{eq:crossterm}
P_{\rm out,\alpha} &= &  M^{\alpha}_{a}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} \nonumber \\
&+& M^{\alpha}_{b}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}
^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&+& 2 M^{\alpha}_{c}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&-& M^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}
Each of these terms is plotted in Figure \ref{fig:sigloss_terms} for two cases: inverse covariance weighting ($\textbf{R}_{r} = \widehat{\textbf{C}}_{r}^{-1}$) and unweighted ($\textbf{R}_{r} = \textbf{I}$). We will now examine the behavior of this equation in the regimes of very weak and very strong injected signal.

{\bf Small injection:}
In this regime, the first and the fourth terms of Equation \eqref{eq:crossterm} are very similar, differing only in their covariance matrices. When the injected signal is very small compared to the data, their difference can be neglected. In this limit we might imagine that the cross-correlation should also be small, enabling us to recover just the EoR term. However, examining the cross-term plotted in Figure \ref{fig:sigloss_terms} (red), we see that this term is still large enough to have a significant 
impact. This is trivially demonstrated in the identity case (right plot) where even at small injects, large cross terms are 
evident.

{\bf Large injection:}
When the injected signal is much larger than the data-only power spectrum, the data-only components can 
be neglected as many orders of magnitude smaller. As we can see from Figure \ref{fig:sigloss_terms}, the cross-terms are also small in comparison. Here only do we recover the ``loss'' injected simulated EoR 
signal as intended (the black and green curves agree).

{\bf In between:}
When the injected signal is of a similar amplitude to the data by itself, the situation becomes less straightforward. We see that 
the weighted injected power spectrum component mirrors the input power indicating little loss (i.e. the green curve follows the dotted black line), eventually 
departing from unity when the injected amplitude is well above the level of the data power spectrum. However, 
the cross-terms have nearly the same amplitude, but with a negative sign. The sum of the two is substantially lower and therefore drives down the $P_{\rm out}$ estimator (black). We note that in A15, signal loss was computed by only looking at the second term in Equation \eqref{eq:crossterm} (green), which incorrectly implies no loss at the data level. Ignoring the effect of the negative power from the cross-terms is the main reason for under-estimating power spectrum limits in A15.

@@

The source of the strong negative cross-term is not immediately obvious, however it is an explainable effect. 
When $\textbf{R}_{r}$
is taken to be $\widehat{\textbf{C}}_{r}^{-1}$, the third term of Equation \eqref{eq:crossterm} is a cross-correlation between $\widehat{\textbf{C}}_{r}^{-1}\textbf{x}$ and
$\widehat{\textbf{C}}_{r}^{-1}\textbf{e}$. As shown in \citet{switzer_et_al2015}, this cross-correlation term is non-zero, and in fact negative in expectation. 
This negative cross-term power arises from a coupling between the inverse of 
$\widehat{\textbf{C}}_{r}$ and $\mathbf{x}$. 
Intuitively, we can see this by expanding the empirical covariance of $\textbf{r}=\textbf{x}+\textbf{e}$ in the
$|\textbf{x}|\ll|\textbf{e}|$ regime:

@@

\noindent where we can neglect the first term because $\textbf{x}$ is small.  Without loss of generality, we will assume
an eigenbasis of $\textbf{e}$, so that $\langle 
\textbf{ee}^{\dagger} \rangle_{t}$ is diagonal. The middle 
two terms, however, can have power in their off-diagonal terms due to the fact that, when averaging over a finite
ensemble, $\langle\textbf{xe}^\dagger\rangle_t$ is not zero.  As shown in Appendix C of \citet{parsons_et_al2014}, to leading order the inversion of a matrix like $\widehat{\textbf{C}}_{r}$ that is diagonal-dominant (from $\langle 
\textbf{ee}^{\dagger} \rangle_{t}$) with smaller
off-diagonal terms results in a new diagonal-dominant matrix with negative off-diagonal terms. These off-diagonal
terms depend on both $\textbf{x}$ and $\textbf{e}$. Then, when $\widehat{\textbf{C}}^{-1}_{r}$ is multiplied into $\textbf{x}$,
the result is a vector that is similar to $\textbf{x}$ but
contains a residual correlation to $\textbf{e}$ from the off-diagonal components of $\widehat{\textbf{C}}^{-1}_{r}$. The
correlation is negative because the product $\widehat{\textbf{C}}_r^{-1}\textbf{x}$ effectively squares the $\textbf{x}$-dependence
of the off-diagonal terms in $\widehat{\textbf{C}}^{-1}_{r}$ while retaining the negative sign that arose from the inversion
of a diagonal-dominant matrix.
% XXX [ARP: we may want to write this all out explicitly in an appendix]

The correlation between $\widehat{\textbf{C}}_r^{-1}\textbf{e}$ and $\widehat{\textbf{C}}_r^{-1}\textbf{x}$ means that we cannot compute signal loss using a signal-only 
simulation, which would yield greater values for $P_{\rm out}$ and thereby underestimate signal loss. This argument is also made in \citet{switzer_et_al2015} in regards to estimating loss associated with the foreground cleaning of Green Bank Telescope (GBT) data. Therefore, in our revised 
signal loss computation we use the full quantity for $P_{\rm out}$ as defined in Equation \eqref{eq:sigloss}, which subtracts the 
weighted power spectrum of the data from the weighted power spectrum of data plus EoR.
